---
title: "Analyses speak smart 2.0"
author: "Thomas Wilschut"
date: "12/1/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

## Analyses

The main goal of the speak smart study is to see if benefits of adaptive word learning, 
are consistently found in typing-based learning, generalize to speech-based learning.
To examine the possibility for speech-based, the speak smart experiment contains 4 learning sessions:
The first session uses the SlimStampen adaptive learning algorithm to determine the repetition schedule 
of the items, and is typing-based (SST). The second session is also typing-based, but uses a less adaptive 
Leitner flashcard system (LS). The third block is speech-based and uses the SlimStampen algorithm (SSS). The
fourth and final block is also speech-based, and uses the Leitner repetition schedule (LS).

# Import packages, load data, and define functions:
```{r, show =FALSE, echo = FALSE}

options(warn=-1)

# packages:
library('tidyverse') 
library('ggplot2') 
library('nlme')
library('plyr') 
library('dplyr') 
library('reshape2')
library('ggridges')
library('BBmisc')
library('quanteda')
library('hrbrthemes')
library('jsonlite')
library('lme4')
library('gridExtra')
library('gghalves')
library('lmerTest') 
library('BayesFactor')
library('ggsignif')

# functions to summarize/show data:
data_summary <- function(data, varname, groupnames){
  require(plyr)
  summary_func <- function(x, col){
    c(mean = mean(x[[col]], na.rm=TRUE),
      se = sd(x[[col]]/ sqrt(length(x[[col]])), na.rm=TRUE))
  }
  data_sum<-ddply(data, groupnames, .fun=summary_func,
                  varname)
  data_sum <- rename(data_sum, c("mean" = varname))
  return(data_sum)
}

GeomSplitViolin <- ggproto("GeomSplitViolin", GeomViolin, 
                           draw_group = function(self, data, ..., draw_quantiles = NULL) {
                             data <- transform(data, xminv = x - violinwidth * (x - xmin), xmaxv = x + violinwidth * (xmax - x))
                             grp <- data[1, "group"]
                             newdata <- plyr::arrange(transform(data, x = if (grp %% 2 == 1) xminv else xmaxv), if (grp %% 2 == 1) y else -y)
                             newdata <- rbind(newdata[1, ], newdata, newdata[nrow(newdata), ], newdata[1, ])
                             newdata[c(1, nrow(newdata) - 1, nrow(newdata)), "x"] <- round(newdata[1, "x"])
                             
                             if (length(draw_quantiles) > 0 & !scales::zero_range(range(data$y))) {
                               stopifnot(all(draw_quantiles >= 0), all(draw_quantiles <=
                                                                         1))
                               quantiles <- ggplot2:::create_quantile_segment_frame(data, draw_quantiles)
                               aesthetics <- data[rep(1, nrow(quantiles)), setdiff(names(data), c("x", "y")), drop = FALSE]
                               aesthetics$alpha <- rep(1, nrow(quantiles))
                               both <- cbind(quantiles, aesthetics)
                               quantile_grob <- GeomPath$draw_panel(both, ...)
                               ggplot2:::ggname("geom_split_violin", grid::grobTree(GeomPolygon$draw_panel(newdata, ...), quantile_grob))
                             }
                             else {
                               ggplot2:::ggname("geom_split_violin", GeomPolygon$draw_panel(newdata, ...))
                             }
                           })

geom_split_violin <- function(mapping = NULL, data = NULL, stat = "ydensity", position = "identity", ..., 
                              draw_quantiles = NULL, trim = TRUE, scale = "area", na.rm = FALSE, 
                              show.legend = NA, inherit.aes = TRUE) {
  layer(data = data, mapping = mapping, stat = stat, geom = GeomSplitViolin, 
        position = position, show.legend = show.legend, inherit.aes = inherit.aes, 
        params = list(trim = trim, scale = scale, draw_quantiles = draw_quantiles, na.rm = na.rm, ...))
}


# import:
dat <- readRDS("data.rds")        # full dataset containing data from all both sessions 
long <- readRDS("data_long.rds")  # data on long-term retention (after 4-8 days)


df = aggregate(correct ~ block + participant_number, data= dat %>% filter (types == 'Test'), sum, na.rm=TRUE)
df2 = aggregate(correct ~ block + participant_number, data= dat %>% filter (types == 'Test'), length)
df$number <- df2$correct
df$prop <- df$correct/df$number
df$method_adaptive <- df$block == 'SST' | df$block == 'SSS'
df$condition <- df$block == 'SSS' | df$block == 'LS'
df[(df$condition == FALSE),'condition'] <- 'Typing'
df[(df$condition == TRUE),'condition'] <- 'Speaking'
df2 <- df %>% filter(method_adaptive == FALSE)
df3 <- df %>% filter(method_adaptive == TRUE)
df2$cor_leitner <- df2$correct
df2$cor_adaptive <- df3$correct

immediate_test <- df2 %>% select(participant_number, cor_leitner, cor_adaptive, condition)
immediate_test$al_benefit <- immediate_test$cor_adaptive - immediate_test$cor_leitner

long_term_retention <- long %>% select(participant_number, cor_leitner, cor_adaptive, condition)
long_term_retention$al_benefit <- long_term_retention$cor_adaptive - long_term_retention$cor_leitner

immediate_test$prop_leitner <- immediate_test$cor_leitner/22
immediate_test$prop_adaptive <- immediate_test$cor_adaptive/22
immediate_test$prop_alb <- immediate_test$prop_adaptive - immediate_test$prop_leitner
immediate_test$overall_prop <- (immediate_test$prop_adaptive + immediate_test$prop_leitner)/2
immediate_test$abl_perc_dif <- -((immediate_test$prop_adaptive/immediate_test$prop_leitner)-1 )*100


long_term_retention$prop_leitner <- long_term_retention$cor_leitner/22
long_term_retention$prop_adaptive <- long_term_retention$cor_adaptive/22
long_term_retention$prop_alb <- long_term_retention$prop_adaptive - long_term_retention$prop_leitner
long_term_retention$overall_prop <- (long_term_retention$prop_adaptive + long_term_retention$prop_leitner)/2
long_term_retention$abl_perc_dif <- as.numeric(((long_term_retention$prop_adaptive/long_term_retention$prop_leitner)-1 )*100)

long_term_retention$abl_perc_dif [!is.finite(long_term_retention$abl_perc_dif )] <- NaN              # convert 'Inf' to NaN
long_term_retention  <- long_term_retention[!is.na(long_term_retention$abl_perc_dif),]  


# Means:
typing_immediate <- immediate_test %>% filter(condition == "Typing")
typing_immediate_sum <- typing_immediate %>% summarise_all(mean)
typing_immediate_sum$cond <- 'Typing, immediate test'

speaking_immediate <- immediate_test %>% filter(condition == "Speaking")
speaking_immediate_sum <- speaking_immediate %>% summarise_all(mean)
speaking_immediate_sum$cond <- 'Speaking, immediate test'

typing_ltr <- long_term_retention %>% filter(condition == "Typing")
typing_ltr_sum <- typing_ltr %>% summarise_all(mean)
typing_ltr_sum$cond <- 'Typing, long term test'

speaking_ltr <- long_term_retention %>% filter(condition == "Speaking")
speaking_ltr_sum <- speaking_ltr %>% summarise_all(mean)
speaking_ltr_sum$cond <- 'Speaking, long term test'

means <- rbind (typing_immediate_sum, speaking_immediate_sum, typing_ltr_sum, speaking_ltr_sum)
means <-  means %>% select (cond, cor_adaptive, cor_leitner, al_benefit, prop_leitner, prop_adaptive, overall_prop, prop_alb, abl_perc_dif)
```



# Plots 1: correctness
```{r, echo=FALSE}
options(warn=-1)

## 1. Correctness during learning phase ##
# prepare:
df <- data_summary(dat %>% filter (types == 'Learning'), varname="correct", 
                    groupnames=c("method_adaptive", "participant_number", "condition"))
df2 <- df %>% filter(method_adaptive == FALSE)
df3 <- df %>% filter(method_adaptive == TRUE)
df2$cor_leitner <- df2$correct
df2$cor_adaptive <- df3$correct

mean_leitner <- mean(df2$cor_leitner)
mean_adaptive <- mean(df2$cor_adaptive)

# plot:
ggplot(df2, aes(x = cor_leitner, y = cor_adaptive, color = condition)) + 
  geom_point(size = 4, na.rm = FALSE, alpha =  0.7) + geom_abline(intercept = 0, slope = 1, colour = 'grey') +   geom_hline(yintercept = mean_adaptive, linetype = "dashed") +
  geom_vline(xintercept = mean_leitner, linetype = "dashed") +
  geom_point(x = mean_leitner, y = mean_adaptive, size = 3, color = "black") +
  coord_fixed() + ylim(0.5,1) + xlim(0.5,1) +
  theme_classic() + scale_color_manual(values=c('#228b22','#006994')) +      
  labs(title="Proportion of correct responses during learning", 
       x="Using the Leitner algortihm", 
       y = "Using the SlimStampen algortihm",
       color = "Learning modality") +
  theme(
    legend.justification = c(1, 0.1), legend.position = c(1, 0.1),
    panel.grid.major = element_line(colour="grey", size = 0.2),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 12, colour = "darkgrey")
  ) 

ggsave("cor_learning.png", width = 5, height = 5, path = "~/Desktop/plots" )

## 2. Correctness during test phase ##
# prepare:
df <- data_summary(dat %>% filter (types == 'Test'), varname="correct", 
                    groupnames=c("method_adaptive", "participant_number", "condition"))
df2 <- df %>% filter(method_adaptive == FALSE)
df3 <- df %>% filter(method_adaptive == TRUE)
df2$cor_leitner <- df2$correct
df2$cor_adaptive <- df3$correct

mean_leitner <- mean(df2$cor_leitner)
mean_adaptive <- mean(df2$cor_adaptive)

# plot:
ggplot(df2, aes(x = cor_leitner, y = cor_adaptive, color = condition)) + 
  geom_point(size = 4, na.rm = FALSE, alpha =  0.7) + geom_abline(intercept = 0, slope = 1, colour = 'grey') +   geom_hline(yintercept = mean_adaptive, linetype = "dashed") +
  geom_vline(xintercept = mean_leitner, linetype = "dashed") +
  geom_point(x = mean_leitner, y = mean_adaptive, size = 3, color = "black") +
  coord_fixed() + ylim(0.5,1) + xlim(0.5,1) +
  theme_classic() + scale_color_manual(values=c('#B22222','#E69F00')) +    
  labs(title="Proportion of correct responses during test", 
       x="Using the Leitner algortihm", 
       y = "Using the SlimStampen algortihm",
       color = "Learning modality") +
  theme(
    legend.justification = c(1, 0.1), legend.position = c(1, 0.1),
    panel.grid.major = element_line(colour="grey", size = 0.2),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 12, colour = "darkgrey")
  ) 

ggsave("cor_test.png", width = 5, height = 5, path = "~/Desktop/plots" )

## 3. Number of items learned, on immediate test ##
# prepare:
df=aggregate(correct ~ block + participant_number, data = dat %>% filter (types == 'Test'), sum, na.rm=TRUE)
df$prop_cor <- df$correct 
df$method_adaptive <- df$block == 'SST' | df$block == 'SSS'
df$condition <- df$block == 'LT'  | df$block ==  'SST'
df2 <- df %>% filter(method_adaptive == FALSE)
df3 <- df %>% filter(method_adaptive == TRUE)
df2$cor_leitner <- df2$prop_cor
df2$cor_adaptive <- df3$prop_cor
df2[(df2$condition == FALSE),'condition'] <- 'Speaking'
df2[(df2$condition == TRUE),'condition'] <- 'Typing'

mean_leitner <- mean(df2$cor_leitner)
mean_adaptive <- mean(df2$cor_adaptive)

# plot:
ggplot(df2, aes(x = cor_leitner, y = cor_adaptive, color = condition)) +
  geom_point(size = 4, na.rm = FALSE, alpha =  0.7,  position = position_jitter()) + 
  geom_hline(yintercept = mean_adaptive, linetype = "dashed") +
  geom_vline(xintercept = mean_leitner, linetype = "dashed") +
    geom_point(x = mean_leitner, y = mean_adaptive, size = 3, color = "black") +
  geom_abline(intercept = 0, slope = 1, colour = 'grey') + coord_fixed() + 
  ylim(0,23) + xlim(0,23) +
  theme_classic() +  scale_color_manual(values=c('#228b22','#006994')) + 
  labs(title="Number of items succesfully recalled immediately", 
       x="After using the Leitner algortihm", 
       y = "After using the SlimStampen algortihm",
       color='Learning modality') +
  theme(
    legend.justification = c(1, 0.1), legend.position = c(1, 0.1),
    panel.grid.major = element_line(colour="grey", size = 0.2),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 12, colour = "darkgrey")
  ) 

ggsave("eff_im.png", width = 5, height = 5, path = "~/Desktop/plots" )

## 4. Number of items learned after 4-8 days ##
# prepare:
mean_leitner <- mean(long$cor_leitner)
mean_adaptive <- mean(long$cor_adaptive)

# plot:
ggplot(long, aes(x = cor_leitner, y = cor_adaptive, color = condition)) +
  geom_point(size = 4, na.rm = FALSE, alpha =  0.7,  position = position_jitter()) + 
  geom_hline(yintercept = mean_adaptive, linetype = "dashed") +
  geom_vline(xintercept = mean_leitner, linetype = "dashed") +
  geom_point(x = mean_leitner, y = mean_adaptive, size = 3, color = "black") +
  geom_abline(intercept = 0, slope = 1, colour = 'grey') + coord_fixed() + 
  ylim(0,23) + xlim(0,23) + # geom_text(aes(label=participant_number),hjust=0, vjust=0) +
  theme_classic() +  scale_color_manual(values=c('#B22222','#E69F00')) + 
  labs(title="Number of items succesfully recalled after 4-8 days", 
       # subtitle = "Based on data from 48 participants, for 2 sessions",
       x="After using the Leitner algortihm", 
       y = "After using the SlimStampen algortihm",
       color='Learning modality') +
  theme(
    legend.justification = c(1, 0.1), legend.position = c(1, 0.1),
    panel.grid.major = element_line(colour="grey", size = 0.2),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 12, colour = "darkgrey")
  ) 


ggsave("eff_lt.png", width = 5, height = 5, path = "~/Desktop/plots" )


## 5. compare typing and speaking 
# prepare:
immediate_test$interval <- 'Immediate'
long_term_retention$interval <- 'Long term'
correctness <- rbind (immediate_test, long_term_retention)

# plot:
ggplot(correctness, aes(condition, overall_prop, fill = interval)) +
  geom_split_violin(alpha = 0.55, color = 'NA') +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.5, show.legend = FALSE) + 
  theme_minimal() + scale_fill_manual(values=c('#B22222', '#E69F00')) +
  labs(title="Learning efficiency for typing- and speech-based learning", 
       x="", 
       y = "Proportion of words successfully recalled",
       fill='Retention Interval'
       ) + 
  theme(
    panel.grid.major = element_line(colour="grey", size = 0.2),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 12, colour = "darkgrey"),
    axis.text.x = element_text(size=12, color = 'black'),
    axis.text.y = element_text(size=11, color = 'black'),
    legend.justification = c(1, 0.1), legend.position = c(1, 0.7),
    legend.background = element_rect(fill="white", size=0.5, color = NA)
  ) 


ggsave("eff_typings_peaking.png", width = 6, height = 6, path = "~/Desktop/plots" )

```

# Plots 2: Response times
```{r, echo=FALSE}
options(warn=-1)
# plot:
ggplot(dat %>% filter (rt < 12000 & rt > 200), aes(block, rt, fill = types)) +
  geom_split_violin(alpha = 0.55, color = 'NA') +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.5, show.legend = FALSE) + 
  theme_minimal() + scale_fill_manual(values=c('#383E42', '#E69F00')) +
  scale_x_discrete(labels = c("SST" = "SS Typing", "SSS" = "SS Speaking", "LS" = "Leitner Speaking", "LT" = "Leitner Typing")) + 
  labs(title="Response times during test and learning", 
       x="", 
       y = "Response time (ms) for correct responses",
       fill='Experiment phase'
       ) + 
  theme(
    panel.grid.major = element_line(colour="grey", size = 0.2),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 12, colour = "darkgrey"),
    axis.text.x = element_text(size=12, color = 'black'),
    axis.text.y = element_text(size=11, color = 'black'),
    legend.justification = c(1, 0.1), legend.position = c(1, 0.7),
    legend.background = element_rect(fill="white", size=0.5, color = NA)
  ) + scale_y_continuous(breaks = seq(0, 6000, length.out = 11), limits = c(0,6000))


ggsave("rts.png", width = 6, height = 6, path = "~/Desktop/plots" )

```

# Plots 3: Effects over time/repetitions
```{r, echo=FALSE, warning=FALSE, message=FALSE}
options(warn=-1)
## 1. Response times over time ##

ggplot(data = dat %>% filter (rt < 12000 & correct & time < 11), aes(time, as.numeric(rt), color = block))  + 
  geom_smooth(formula = y ~ x, method = "loess", span = 0.5) + 
  theme_classic() + scale_color_manual(values=c('#228b22','#006994', '#383E42', '#E69F00')) + 
  labs(title="RTs over learning time", 
       subtitle = "For correct responses only",
       x="Learning time (minutes)", 
       y = "Response time (ms)",
       color='Learning condition') +
  theme(
    legend.justification = c(1, 0.1), legend.position = c(1, 0.6),
    panel.grid.major = element_line(colour="grey", size = 0.2),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 12, colour = "darkgrey")
  ) 


ggsave("rt_over_time.png", width = 8, height = 4, path = "~/Desktop/plots" )

## Correctness over repetitions
#prepare:

df_total <- dat
df_total$repetition <- df_total$repetition - 1
df_total <- df_total[!is.na(df_total$correct),]  
df_total <- df_total[!is.na(df_total$repetition),]  
df_total <- df_total[!is.na(df_total$block),]  



df_sum <- df_total %>% filter(repetition <=11) %>%
  group_by(block, repetition) %>%
  summarise_at(vars(correct), list(name = mean))

ggplot(df_sum, aes(repetition, name, color = block)) + geom_line() + geom_point() + 
theme_classic() + scale_color_manual(values=c('#228b22','#006994', '#383E42', '#E69F00')) + geom_point(alpha=0.01) +
  labs(title="Correctness over repetitions", 
       x="Item repetitions", 
       y = "Proportion of correct responses",
       color='Learning condition') +
  theme(
    legend.justification = c(1, 0.1), legend.position = c(1, 0.05),
    panel.grid.major = element_line(colour="grey", size = 0.2),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 12, colour = "darkgrey")
  ) + scale_x_continuous(breaks = seq(0, 12, length.out = 13), limits = c(0,12)) + ylim(0,1)

ggsave("cor_over_reps.png", width = 8, height = 4, path = "~/Desktop/plots" )
```

# Descriptives 

```{r}
# number of trials:
summary <- aggregate(id ~ block + types + participant_number, data = dat, FUN = length)
trial <- aggregate(id ~ block + types, data = summary, FUN = mean)
trialsd <- aggregate(id ~ block + types, data = summary, FUN = sd)

# correctness:
summary <- aggregate(correct ~ block + types + participant_number, data = dat, FUN = mean)
cor <- aggregate(correct ~ block + types, data = summary, FUN = mean)
corsd <- aggregate(correct ~ block + types, data = summary, FUN = sd)

# rts
rtclean <- dat
rtclean$rt[!is.finite(rtclean$rt)] <- NaN              # convert 'Inf' to NaN
rtclean <- rtclean[!is.na(rtclean$rt),]  

summary <- aggregate(rt ~ block + types + participant_number, data = rtclean, FUN = mean, na.omit = TRUE)
rts <- aggregate(rt ~ block + types, data = summary, FUN = mean)
rtsd <- aggregate(rt ~ block + types, data = summary, FUN = sd)

descr <- rts

descr$rt_sd <- rtsd$rt
descr$cor <- cor$correct
descr$cor_sd <- corsd$correct
descr$trial <- trial$id
descr$trialsd <- trialsd$id

descr
means

```

# Linear Mixed Effects models for performance during test
```{r, echo = FALSE}

# recode session to contrast coding:
dat[(dat$session == 1),'session'] <- -0.5   
dat[(dat$session == 2),'session'] <- 0.5
rtclean[(rtclean$session == 1),'session'] <- -0.5   
rtclean[(rtclean$session == 2),'session'] <- 0.5

# RT on test for speaking, for correct responses:
rt_speaking <- lmer(rt~ method_adaptive + session + ((1|participant_number) + (1|id)), 
                    data = rtclean %>% filter(condition == 'Speaking' & types == 'Test' & correct == TRUE))
summary(rt_speaking)

# RT on test for typing, for correct responses:
rt_typing <- lmer(rt~ method_adaptive + session + ((1|participant_number) + (1|id)),
                  data = rtclean %>% filter(condition == 'Typing' & types == 'Test'  & correct == TRUE))
summary(rt_typing)

# Correctness on test for speaking:
cor_speaking <- glmer(correct ~ method_adaptive + session + ((1 | participant_number) + ( 1 | id)), 
                    data = dat %>% filter (condition == 'Speaking' & types == 'Test'), family = 'binomial')
summary(cor_speaking)

# Correctness on test for typing:
cor_typing <- glmer(correct ~ method_adaptive + session + ((1 | participant_number) + ( 1 | id)), 
                    data = dat %>% filter (condition == 'Typing' & types == 'Test'), family = 'binomial')
summary(cor_typing)
```
# Comparing ALBs and efficiency 

``` {r}

# SS vs Leitner, immeidate test:
ttestBF(x = immediate_test$prop_leitner, y = immediate_test$prop_adaptive, paired=TRUE)

# SS vs Leitner, long-term:
ttestBF(x = long_term_retention$prop_leitner, y = long_term_retention$prop_adaptive, paired=TRUE)

# Immediate test typing versus speaking:
im <- full_join(typing_immediate , speaking_immediate, by = 'participant_number')
im <- na.omit(im)
ttestBF(x = im$overall_prop.x, y = im$overall_prop.y, paired=TRUE, na.rm = TRUE)

# LT typing versus speaking:
ttestBF(x = typing_ltr$overall_prop, y = speaking_ltr$overall_prop, paired=FALSE)  # here, a paired test is not possible 

```
